---
title: "The Anatomy of a Plot"
author: "Daniel Marcelino"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    number_sections: yes
    toc_depth: 2
    toc: yes
    fig_width: 5
vignette: >
  %\VignetteIndexEntry{The Anatomy of a Plot}
   %\VignetteEngine{knitr::rmarkdown}
    \usepackage[T1]{fontenc}
     \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, echo=TRUE, message=FALSE, comment=NA, fig.align="center", fig.width=5, fig.height=3.5)
```
  

# Overview
This document presents several wrapper functions to deal with a ggplot2 object.



# Preview ggplot2 themes
Used to preview ggplot2 themes.
```{r}
library("SciencesPo", quietly = TRUE)

Graph()
```


# Themes

```{r}
Graph() + theme_darkside()
```


# Title 

## Align title right
Align ggplot2 title to the right.
```{r}
Graph() + align_title_right()
```


## Align title left 
Align ggplot2 title to the left.
```{r}
Graph() + align_title_left()
```

## Remove ggplot2 title
Do not show plot title.
```{r}
Graph() + no_title()
```


# Plot Axes

## Remove ggplot2 x-axis
Remove the x axis from a ggplot2 including the axis lines, titles, labels and ticks.

```{r}
Graph() + no_x_axis()
```


## Remove ggplot2 y-axis
Remove the y axis from a ggplot2 including the axis lines, titles, labels and ticks.

```{r}
Graph() + no_y_axis()
```


## Remove ggplot2 Axes
Remove both axes from a ggplot2 including the axes lines, titles, labels and ticks.

```{r}
Graph() + no_axes()
```


## Remove ggplot2 x-axis line
Do not show x-axis line.

```{r}
Graph() + no_x_line()
```


## Remove ggplot2 y-axis line
Do not show y-axis line.
```{r}
Graph() + no_x_line()
```


## Remove ggplot2 x-axis text/labels
Do not show x-axis text/labels
```{r}
Graph() + no_x_text()
```


## Remove ggplot2 y-axis text/labels
Do not show y-axis text/labels.
```{r}
Graph() + no_y_text()
```


## Remove ggplot2 x-axis tick marks
Do not show tick marks for x-axis.
```{r}
Graph() + no_x_ticks()
```

## Remove ggplot2 y-axis tick marks
Do not show tick marks for y-axis.
```{r}
Graph() + no_y_ticks()
```


## Remove ggplot2 tick marks
Do not show tick marks for all axes.

```{r}
Graph() + no_ticks()
```


## Remove ggplot2 x-axis titles
Do not show x-axis title.
```{r}
Graph() + no_x_title()
```


## Remove ggplot2 y-axis titles
Do not show y-axis title.
```{r}
Graph() + no_y_title()
```


## Remove ggplot2 axes titles
Do not show titles for all axes.
```{r}
Graph() + no_axes_titles()
```


## Rotate ggplot2 x-axis labels
Rotate the x-axis labels/text by a given angle.
```{r}
Graph() + rotate_x_text(90)
```


## Rotate ggplot2 y-axis labels
Rotate the y-axis labels/text by a given angle.
```{r}
Graph() + rotate_y_text(65)
```


## Rotate ggplot2 axes labels
Rotate both axes labels by a given angle.
```{r}
Graph() + rotate_axes_text(45)
```


# Plot Gridlines

## Remove vertical gridlines
Remove all major and minor vertical gridlines from a ggplot2 plot or theme.
```{r}
Graph() + no_x_gridlines()
```


## Remove horizontal gridlines
Remove all major and minor horizontal gridlines from a ggplot2 plot or theme.
```{r}
Graph() + no_y_gridlines()
```


## Remove all gridlines
Remove all vertical and horizontal gridlines (major and minor) from a ggplot2 plot or theme.
```{r}
Graph() + no_gridlines()
```


## Remove major x gridlines
Do not show major gridlines for x-axis.
```{r}
Graph() + no_major_x_gridlines()
```


## Remove major y gridlines
Do not show major gridlines for y-axis.
```{r}
Graph() + no_major_y_gridlines()
```


## Remove major gridlines
Do not show major gridlines.
```{r}
Graph() + no_major_gridlines()
```


## Remove minor vertical gridlines
Do not show minor gridlines for x-axis.
```{r}
Graph() + no_minor_x_gridlines()
```


## Remove minor horizontal gridlines
Do not show minor gridlines for y-axis.
```{r}
Graph() + no_minor_y_gridlines()
```


## Remove minor gridlines
Do not show minor gridlines.
```{r}
Graph() + no_minor_gridlines()
```


# Legend

## Remove ggplot2 legend
Remove the legend from a ggplot2 plot.
```{r}
Graph() + no_legend()
```


## Remove ggplot2 legend title
Remove the legend title from a ggplot2 plot.
```{r}
Graph() + no_legend_title()
```

## Legend Position
Move legend to given location. The location of the legend is 'top', 'bottom', 'right', 'left' or 'none'.

```{r}
Graph() + move_legend('bottom')
```

## Legend to the bottom
Move a ggplot2 legend to the bottom of the plot.
```{r}
Graph() + legend_bottom()
```


## Legend to the top
Move a ggplot2 legend to the top of the plot.
```{r}
Graph() + legend_top()
```


## Legend to the left
Move a ggplot2 legend to the left of the plot.
```{r}
Graph() + legend_left()
```


## Legend to the right
Move a ggplot2 legend to the right of the plot.
```{r}
Graph() + legend_right()
```
