---
title: "Calculating Probability of a binary outcome using Binary Logistic"
author: "Daniel"
date: "July 17, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




## First SlideProject Description

    A generalized linear model that to estimate the probability of a vehicle having a V-engine or a straight engine using the user-defined a weight (lbs) and engine displacement values (cu. inches).
    Generalized linear or regression is useful when you are predicting a binary outcome (i.e. V-engine or straight engine in this example) from a set of continuous predictor variables.
    The data for this project (mtcars) was published in the 1974 Motor Trend US magazine, and is comprised of fuel consumption, performance and physical features for 32 automobiles.



## Data
```{r}

```


## generalized linear model


```{r}

## Making probability prediction

Our goal here is to calculate a predicted probability of a V engine, for specific values of the predictors:

    a weight of 2100 lbs and
    engine displacement of 180 cubic inches.

  wt = 1.2
  disp = 180 
  data(mtcars)
  model = glm(formula= vs ~ wt + disp, data=mtcars, family=binomial)
  newdata = data.frame(wt = wt, disp = disp)
  prob = predict(model, newdata, type="response")

The predicted probability is 0.07

```

