---
title: "Measures of Political Behavior"
author: "Daniel Marcelino"
date: "`r Sys.Date()`"
output: pdf_document
 # rmarkdown::html_vignette:
  #  fig_width: 5
vignette: >
  %\VignetteIndexEntry{Measures of Political Behavior}
   %\VignetteEngine{knitr::rmarkdown}
    \usepackage[T1]{fontenc}
     \usepackage[utf8]{inputenc}
     %\usepackage[portuguese]{babel}
---
  
## Introduction

The very objective of the *SciencesPo* package is to provide methods for computing some of the popular measures used in political science literature, such as the indices of political concentration--or fragmentation, inequality, and seat apportionment methods. The following examples demontrates how to use some of the applications available in the R library *SciencesPo*.

## Indices of Political Diversity 

```{r politicalDiversity1, echo=TRUE, message=FALSE, comment=NA}
library("SciencesPo")

# The 2004 presidential election in the US (vote share):

US2004 <- c("Democratic"=0.481, "Republican"=0.509, 
            "Independent"=0.0038, "Libertarian"=0.0032, 
            "Constitution"=0.0012, "Green"=0.00096,
            "Others"=0.00084)

print(US2004)

politicalDiversity(US2004); # ENEP (laakso/taagepera) method 

politicalDiversity(US2004, index= "golosov");

politicalDiversity(US2004, index= "herfindahl");
```

Considers the following `data.frame` with electoral results for the 1999 election in Helsinki, the seats were allocated using both the Saint-Laguë and the D'Hondt methods: 

```{r Helsinki-election, echo=TRUE, message=FALSE}
# Helsinki's 1999

Helsinki <- data.frame(
  votes = c(68885, 18343, 86448, 21982, 51587,
            27227, 8482, 7250, 365, 2734, 1925,
            475, 1693, 693, 308, 980, 560, 590, 185),
  seats.SL=c(5, 1, 6, 1, 4, 2, 1, 0, 0, 0, 0, 0, 0, 0,
             0, 0, 0, 0, 0),
  seats.dH=c(5, 1, 7, 1, 4, 2, 0, 0, 0, 0, 0, 0, 0, 0,
             0, 0, 0, 0, 0))
```

```{r politicalDiversity2, echo=TRUE, message=FALSE, comment=NA}
# politicalDiversity(Helsinki$votes); #ENEP Votes

politicalDiversity(Helsinki$seats.SL); #ENP for Saint-Lague

politicalDiversity(Helsinki$seats.dH); #ENP for D'Hondt
```

Most democratic countries use apportionment methods to transform election results into whole numbers, which indicate the number of seats that each party obtained in a legislative body. Which apportionment method does this best is not a trivial topic in political science and that several methods have been proposed. The following sections briefly present some of these apportionment methods.

## Highest Averages Methods of Allocating Seats Proportionally

Highest averages methods allocate seats proportionally to the number of votes by assigning seats in a way that assures the highest quotient by seat for each party. In what follows are illustrative examples of the operation of some highest averages formulas in a six member district with four parties taken from Lijphart (1994).

```{r table_A.1, echo=TRUE, cache=TRUE}
# Table A.l
lijphart <- c("A"=41000, "B"=29000,"C"=17000, "D"=13000)
```

The basic imputs for this class of functions are: 1) a list of parties, 2) a list of positive votes, and 3) a constant value for the number of seats to be returned. A numeric value (0~1) for the threshold is optional.

### d'Hondt
Under the D'Hondt formula, seats are allocated using divisors of 1, 2, 3, 4, etc. Thus, suggesting that the divisors are simply one more than the number of seats that party already has. 

```{r highestAverages1, message=FALSE, comment=NA}
# The d'Hondt will give the same results as the Jefferson's method (jef).
highestAverages(parties=names(lijphart), votes=lijphart,
                seats = 6, method = "dh") 
```

The d'Hondt is only one way of allocating seats in party list systems. Other methods include the Saint-Laguë, the modified Saint-Laguë, the Danish version, Imperiali (do not to confuse with the Imperiali quota, which is a Largest remainder method), Hungtinton-Hill, Webster, etc. 


### Saint-Laguë

Named after its founder, the pure Sainte-Laguë formula, also known as Schepers method, uses the odd-integer divisor series (1, 3, 5, 7, ...) to elect candidates from political parties in approximate proportion as the proportion of votes won by the party. The outcome would be the same as the outcome produced by the Webster's divisors method.

```{r highestAverages2, echo=TRUE, message=FALSE, comment=NA}
# The Sainte-Laguë will give the same results as the Webster's method (wb)
highestAverages(parties=names(lijphart), votes=lijphart,
                seats = 6, method = "sl") 
```

### Modified Saint-Laguë

The modified Sainte-Laguë use divisors of 1.4, 3, 5, 7, etc. How does the d'Hondt divisor compare to the Modified Sainte-Laguë one? Which divisor method tend to favor larger parties and which one smaller parties?

```{r highestAverages3, echo=TRUE, message=FALSE, comment=NA}
# Seats allocated using modified Sainte-Lague divisors of 1.4, 3, 5.
highestAverages(parties=names(lijphart), votes=lijphart, 
                seats = 6, method = "msl") 
```


### Danish

The Danish divisors (1, 4, 7, 10, ...) increase so fast that large parties are quickly cut down, acting much to the benefit of smaller parties.

```{r highestAverages4, echo=TRUE, message=FALSE, comment=NA}
highestAverages(parties=names(lijphart), votes=lijphart, 
                seats = 6, method = "danish") 
```

### Hungarian

The Hungarian version of Sainte-Laguë uses divisors of (1.5, 3, 5, 7, 9, ...) increase so fast that large parties are quickly cut down, acting much to the benefit of smaller parties.

```{r highestAverages5, echo=TRUE, message=FALSE, comment=NA}
highestAverages(parties=names(lijphart), votes=lijphart, 
                seats = 6, method = "danish") 
```


### Webster

This approach does not use a standard divisor either. It uses instead a different divisor and modified quota such that rounding with cut-off point at *algebraic* mean of lower and upper quotas: wb = $\frac{L + (L + 1)}{2}$.

```{r highestAverages6, echo=TRUE, message=FALSE, comment=NA}
highestAverages(parties=names(lijphart), votes=lijphart, 
                seats = 6, method = "wb") 
```


### Imperiali

The Italian Imperiali divisor system has a slowly-increasing sequence (1, 1.5, 2, 2.5, 3, ...), with 0.5 difference between consecutive divisors. This is tighter than the d'Hondt divisors for which this difference is 1. Like d'Hondt, the Imperiali system is designed to encourage coalitions and secure majority governments under a PR system, but the Imperiali will be the most favourable to large parties as they will gain many seats before their quotients are reduced below those of the smaller parties.

```{r highestAverages7, echo=TRUE, message=FALSE, comment=NA}
highestAverages(parties=names(lijphart), votes=lijphart, 
                seats = 6, method = "imperiali") 
```

The Belgian municipal councils were elected by the Imperiali highest averages method. The following example comes from the 2006 election in Bruges, where 47 seats were upon.

```{r highestAverages8, echo=TRUE, message=FALSE, comment=NA}
Bruges=c("CD&V/N-VA"=32092, "SP.A/Spirit"=20028, 
         "Flemish Interest"=13408, "Open VLD/Vivant"=9520,
         "Green!"=5328, "Other"=2207)

highestAverages(parties=names(Bruges), votes=Bruges, 
                seats = 47, method = "imperiali") 
```

### Hungtinton-Hill

The method proposed by Joseph Hill and Edward Huntington does not use a standard divisor, but a different one and also a modified quota such that rounding with cut-off point at *geometric* mean: hh = $\sqrt{L(L + 1)}$. This method makes sense only if every party is guaranteed at least one seat. It has been used for allotting seats in the US House of Representatives to the states.


```{r highestAverages9, echo=TRUE, message=FALSE, comment=NA}
highestAverages(parties=names(lijphart), votes=lijphart, 
                seats = 6, method = "hh") 
```

### Using Thresholds

Let's assume that we have an election with 1,000 total voters in which five parties (A, B, C, D, and E) have gained 100 (10\%), 150 (15\%), 300 (30\%), 400 (40\%), and 50 (5\%) votes, respectively. In this electoral constituency, there are 3 seats up for election, and all votes cast are valid. The electoral system has a 7\% vote threshold, meaning that parties must get at least 7\% of the total unspoiled votes cast in order to participate in the distribution of seats. Party *E* would then be elimiated from competition at the outset. 

If the d'Hondt method of seat allocation were employed in this hypothetical election, then party *C* would get 1 seat (or 33\% of the number of seats), and party *D* 2 seats (or 67\% of the seats).

```{r highestAverages10, echo=TRUE, message=FALSE, comment=NA}

const <- c("A"=100, "B"=150,"C"=300, "D"=400, "E"=50)

highestAverages(parties=names(const), votes=const,
               seats = 3, method = "dh", threshold = 7/100) 
```

Other methods divide the votes by a mathematically derived quota, such as the Droop quota, the Hare quota (or Hamilton/Vinton), or the Imperiali quota, see next. 


## Largest Remainder Methods of Allocating Seats Proportionally

Highest averages methods allocate seats proportionally to the number of votes by dividing the number of votes required for each party by a quota. The quota may vary because it is the result of dividing the number of unspoiled votes ($v$) by some whole number close to the number of seats that would next be assigned ($s$). In what follows are llustrative examples of the operation of some largest remainders formulas in an eight member district with the same four parties of the previous examples.

### Hare quota
The Droop quota is obtained by quota = $\frac{v}{(s)}$.

```{r largestRemainders1, eval=FALSE, echo=TRUE, message=FALSE, comment=NA}
largestRemainders(parties=names(lijphart), votes=lijphart, 
                seats = 8, method = "hare") 
```

### Droop quota
The Droop quota is obtained by quota = $\frac{v}{(s + 1)}+1$.

```{r largestRemainders2, eval=FALSE, echo=TRUE, message=FALSE, comment=NA}
largestRemainders(parties=names(lijphart), votes=lijphart, 
                seats = 8, method = "droop") 
```


### Hagenbach-Bischoff
The Hagenbach-Bischoff quota is obtained by $\frac{v}{(s + 1)}$.

```{r largestRemainders3, eval=FALSE, echo=TRUE, message=FALSE, comment=NA}
largestRemainders(parties=names(lijphart), votes=lijphart, 
                seats = 8, method = "hagb") 
```


### Imperiali quota
The Imperiali quota is obtained by $\frac{v}{(s + 2)}$. For Italian elections during the 1950s, the quota used was a reinforced version of the quota: $\frac{v}{(s + 3)}$.
 
```{r data-Italy, eval=FALSE, echo=TRUE, message=FALSE}
# The 1946 Italian Constituent Assembly election results: parties and unspoilt votes

Italy = data.frame(party=c("DC", "PSIUP", "PCI", "UDN", "UQ", "PRI",
                            "BNL", "PdA", "MIS", "PCd'I", "CDR",
                           "PSd'Az", "MUI", "PCS", "PDL", "FDPR"),
                   votes=c(8101004, 4758129, 4356686, 1560638,	1211956,
                           1003007, 637328, 334748, 171201, 102393,
                           97690, 78554, 71021, 51088, 40633, 21853))
```

```{r largestRemainders4, eval=FALSE, echo=TRUE, message=FALSE, comment=NA}
with(Italy, largestRemainders(parties=party, votes=votes, 
                seats = 556, method = "imperiali.q") )
```


### Suitable output for recycling within RMarkdown documents 

The output produced by `highestAveragesof()` and `largestRemainders()`  functions is a `data.frame`. Thus, it should be very straightforward to use with other aplications. For instance, using the output with the **knitr** package to produce publishable-quality tables, or graphs with **ggplot2**.

Let's take the data from 2014 Brazilian legislative elections, especifically from one electoral district--the state of Ceará to compare how the caucus would look like among different apportionment methods.

```{r Ceara-election, echo=TRUE, cache=TRUE}
# Results for the state legislative house of Ceara (2014):
Ceara <- c("PCdoB"=187906, "PDT"=326841,"PEN"=132531, "PMDB"=981096,
           "PRB"=2043217,"PSB"=15061, "PSC"=103679, "PSTU"=109830,
           "PTdoB"=213988, "PTC"=67145, "PTN"=278267)
```


Make a table with the d'Hondt formula results:

```{r highestAverages11, echo=TRUE, message=FALSE, comment=NA}
mytable = highestAverages(parties=names(Ceara), votes=Ceara, 
                seats = 42, method = "dh") 

library(knitr)
kable(mytable, align=c("l","c","c"), caption="Election Outcome: d'Hondt")
```


Next, we produce a graph comparing three highest-averages formulas.

```{r largestRemainders5, eval=TRUE, echo=TRUE, message=FALSE, fig.width=7, fig.height=4.5, fig.align="center", fig.cap= "2014 Legislative Election in Ceara (M=42)"}

out1 = highestAverages(parties=names(Ceara), votes=Ceara, 
                seats = 42, method = "dh")
out2 = highestAverages(parties=names(Ceara), votes=Ceara, 
                seats = 42, method = "imperiali") 
out3 = highestAverages(parties=names(Ceara), votes=Ceara, 
                seats = 42, method = "sl")

# add the method:
out1$Method = "d'Hondt"
out2$Method = "imperiali"
out3$Method = "Saint-Laguë"


data <- rbind(out1, out2, out3)

p = ggplot(data=data, aes(x=reorder(Party, -Seats), y=Seats, fill=Method)) +
    geom_bar(stat="identity",position=position_dodge()) +
   labs(x="", y="Seats")
p + scale_fill_fte() + 
  theme_fte(legend = "top") 
```


```{r largestRemainders6, eval=TRUE, echo=TRUE, message=FALSE, fig.width=7, fig.height=4.5, fig.align="center", fig.cap= "2014 Legislative Election in Ceara (M=42)"}

```



##  Measures of Proportionality

Let's consider the following two real-data examples taken from the Wikipedia. These two elections refer to subnational elections, one from Queensland (AUS) and the other from Quebec (CAN).  

```{r Queensland-election, echo=TRUE, cache=TRUE}
# 2012 Queensland state elecion:
Queensland <- data.frame(party = c("LNP", "ALP", "Katter", "Greens", "Ind", "Others"),
                         votes = c(1214553,652092,282098,184147,77282,35794),
                         pvotes = c(49.65, 26.66, 11.5, 7.53, 3.16, 1.47),
                         seats = c(78, 7, 2, 0, 2, 0),
                         pseats = c(87.64, 7.87, 2.25, 0.00, 2.25, 0.00))
```


```{r Quebec-election, echo=TRUE, cache=TRUE}
# 2012 Quebec provincial election:
Quebec <- data.frame(party = c("PQ", "Lib", "CAQ", "QS", "Option", "Other"),
                         pvotes = c(31.95, 31.20, 27.05, 6.03, 1.89, 1.88),
                         pseats = c(54, 50, 19, 2, 0, 0))
```


### The Gallagher Index

```{r proportionality1, echo=TRUE, message=FALSE, comment=NA}
with(Queensland, gallagher(pvotes, pseats))

with(Quebec, gallagher(pvotes, pseats))
```


### The Lijphart's Index of Proportionality

```{r proportionality2, echo=TRUE, message=FALSE, comment=NA}
with(Queensland, lijphart(pvotes, pseats))

with(Quebec, lijphart(pvotes, pseats))
```


### The Grofman Index

```{r proportionality3, echo=TRUE, message=FALSE, comment=NA}
with(Queensland, grofman(pvotes, pseats))

with(Quebec, grofman(pvotes, pseats))
```


### Farina Index

```{r proportionality4, echo=TRUE, message=FALSE, comment=NA}
with(Queensland, farina(pvotes, pseats))

with(Quebec, farina(pvotes, pseats))
```


### The Cox-Shugart measure of proportionality

```{r proportionality5, echo=TRUE, message=FALSE, comment=NA}
with(Queensland, cox.shugart(pvotes, pseats))

with(Quebec, cox.shugart(pvotes, pseats))
```


### The inverse Cox-Shugart measure of proportionality

```{r proportionality6, echo=TRUE, message=FALSE, comment=NA}
with(Queensland, inv.cox.shugart(pvotes, pseats))

with(Quebec, inv.cox.shugart(pvotes, pseats))
```


## Measures of Inequality and Concentration
I will write soon... 



## References

Cowell, F. A. (2000) Measurement of Inequality in Atkinson, A. B. / Bourguignon, F. (Eds): Handbook of Income Distribution. Amsterdam.

Cowell, F. A. (1995) Measuring Inequality Harvester Wheatshef: Prentice Hall.

Benoit, Kenneth. (2000) "Which Electoral Formula is the Most Proportional? A New Look with New Evidence." Political Analysis 8(4): 381-388.

Blais, A., and L. Massicotte (1997) Electoral Formulas: A Macroscopic Perspective, European Journal of Political Research, 32, 107–129.

Gallagher, Michael and Paul Mitchell (2005) The Politics of Electoral Systems. Oxford University Press.

Gallagher, Michael (1992) "Comparing Proportional Representation Electoral Systems: Quotas, Thresholds, Paradoxes and Majorities". British Journal of Political Science, 22, 4, 469-496.

Gallagher, M. (1991) Proportionality, disproportionality and electoral systems. Electoral Studies 10(1):33-51.

Golosov, Grigorii (2010) The Effective Number of Parties: A New Approach, Party Politics, 16: 171-192.

Laakso, Markku and Rein Taagepera (1979) Effective Number of Parties: A Measure with Application to West Europe, Comparative Political Studies, 12: 3-27.
 
Lijphart, Arend (1994) Electoral Systems and Party Systems: A Study of Twenty-Seven Democracies, 1945-1990. Oxford University Press.

Nicolau, Jairo (2008) Sistemas Eleitorais. Rio de Janeiro, FGV.

Norris, Pippa (1997) Choosing electoral systems: proportional, majoritarian and mixed systems.

Taagepera, R., and B. Grofman (2003) Mapping the indices of seats-votes disproportionality and inter-election volatility. Party Politics 9, 6: 659-77.

Taagepera, Rein and Matthew S. Shugart (1989) Seats and Votes: The Effects and Determinants of Electoral Systems. New Haven: Yale University Press.
