---
title: "The Anatomy of a Plot"
author: "Daniel Marcelino"
date: "`r as.character(Sys.Date())`"
output:
  rmarkdown::html_vignette:
    number_sections: yes
    toc_depth: 2
    toc: yes
    fig_width: 5
vignette: >
  %\VignetteIndexEntry{The Anatomy of a Plot}
   %\VignetteEngine{knitr::rmarkdown}
    \usepackage[T1]{fontenc}
     \usepackage[utf8]{inputenc}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, echo=TRUE, message=FALSE, comment=NA, fig.align="center", fig.width=5, fig.height=4, dpi = 72, fig.show = "hold", fig.keep="last", sanitize=TRUE)
```


# Overview
This document presents several wrapper functions to make plotting with ggplot2 a lot easier. 


# Preview ggplot2 themes
`Previewplot()` can be used to preview ggplot2 themes and features.
```{r}
library("SciencesPo", quietly = TRUE)

Previewplot()
```


# Themes

```{r}
Previewplot() + theme_pub()
```


```{r}
Previewplot() + theme_darkside()
```


```{r, eval=FALSE, echo=FALSE, include=FALSE}
x <- rnorm(20000)
y <- x + rnorm(20000, 0.05)
df <- data.frame(x = x, y = y,
  d = densCols(x, y, colramp = colorRampPalette(rev(rainbow(10, end = 4/6)))))
gg <- ggplot(df) +
    geom_point(aes(x, y, col = d), size = 1) +
    scale_color_identity()
print(gg)
```


```{r, echo=FALSE}
library("deldir", quietly = TRUE)

set.seed(105)
long<-rnorm(20,-98,15)
lat<-rnorm(20,39,10)
df <- data.frame(lat,long)

voronoi <- deldir(df$long, df$lat)

gg <- ggplot(data=df, aes(x=long,y=lat)) +
  geom_segment(
    aes(x = x1, y = y1, xend = x2, yend = y2),
    size = 2,
    # Delaunay triangulation
     # voronoi$delsgs
    data = voronoi$dirsgs,
    linetype = 1,
    color= "#FFB958")
gg <- gg + geom_point(
    fill=rgb(70,130,180,255,maxColorValue=255),
    pch=21,
    size = 4,
    color="#333333")
gg <- gg + labs(x = 'lat', y = 'long', title = "My Voronoi Line Segments")
```

# Title 

## Align title right
Align ggplot2 title to the right.
```{r}
gg + align_title_right()
```


## Align title left 
Align ggplot2 title to the left.
```{r}
gg + align_title_left()
```

## Remove ggplot2 title
Do not show plot title.
```{r}
gg + no_title()
```


# Plot Axes

## Remove ggplot2 x-axis
Remove the x axis from a ggplot2 including the axis lines, titles, labels and ticks.

```{r}
gg + no_x_axis()
```


## Remove ggplot2 y-axis
Remove the y axis from a ggplot2 including the axis lines, titles, labels and ticks.

```{r}
gg + no_y_axis()
```


## Remove ggplot2 Axes
Remove both axes from a ggplot2 including the axes lines, titles, labels and ticks.

```{r}
gg + no_axes()
```


## Remove ggplot2 x-axis line
Do not show x-axis line.

```{r}
gg + no_x_line()
```


## Remove ggplot2 y-axis line
Do not show y-axis line.
```{r}
gg + no_x_line()
```


## Remove ggplot2 x-axis text/labels
Do not show x-axis text/labels
```{r}
gg + no_x_text()
```


## Remove ggplot2 y-axis text/labels
Do not show y-axis text/labels.
```{r}
gg + no_y_text()
```


## Remove ggplot2 x-axis tick marks
Do not show tick marks for x-axis.
```{r}
gg + no_x_ticks()
```

## Remove ggplot2 y-axis tick marks
Do not show tick marks for y-axis.
```{r}
gg + no_y_ticks()
```


## Remove ggplot2 tick marks
Do not show tick marks for all axes.

```{r}
gg + no_ticks()
```


## Remove ggplot2 x-axis titles
Do not show x-axis title.
```{r}
gg + no_x_title()
```


## Remove ggplot2 y-axis titles
Do not show y-axis title.
```{r}
gg + no_y_title()
```


## Remove ggplot2 axes titles
Do not show titles for all axes.
```{r}
gg + no_axes_titles()
```


## Rotate ggplot2 x-axis labels
Rotate the x-axis labels/text by a given angle.
```{r}
gg + rotate_x_text(90)
```


## Rotate ggplot2 y-axis labels
Rotate the y-axis labels/text by a given angle.
```{r}
gg + rotate_y_text(65)
```


## Rotate ggplot2 axes labels
Rotate both axes labels by a given angle.
```{r}
gg + rotate_axes_text(45)
```


# Plot Gridlines

## Remove vertical gridlines
Remove all major and minor vertical gridlines from a ggplot2 plot or theme.
```{r}
gg + no_x_gridlines()
```


## Remove horizontal gridlines
Remove all major and minor horizontal gridlines from a ggplot2 plot or theme.
```{r}
gg + no_y_gridlines()
```


## Remove all gridlines
Remove all vertical and horizontal gridlines (major and minor) from a ggplot2 plot or theme.
```{r}
gg + no_gridlines()
```


## Remove major x gridlines
Do not show major gridlines for x-axis.
```{r}
gg + no_major_x_gridlines()
```


## Remove major y gridlines
Do not show major gridlines for y-axis.
```{r}
gg + no_major_y_gridlines()
```


## Remove major gridlines
Do not show major gridlines.
```{r}
gg + no_major_gridlines()
```


## Remove minor vertical gridlines
Do not show minor gridlines for x-axis.
```{r}
gg + no_minor_x_gridlines()
```


## Remove minor horizontal gridlines
Do not show minor gridlines for y-axis.
```{r}
gg + no_minor_y_gridlines()
```


## Remove minor gridlines
Do not show minor gridlines.
```{r}
gg + no_minor_gridlines()
```


# Legend

## Remove ggplot2 legend
Remove the legend from a ggplot2 plot.
```{r}
gg + no_legend()
```


## Remove ggplot2 legend title
Remove the legend title from a ggplot2 plot.
```{r}
gg + no_legend_title()
```

## Legend Position
Move legend to given location. The location of the legend is 'top', 'bottom', 'right', 'left' or 'none'.

```{r}
gg + move_legend('bottom')
```

## Legend to the bottom
Move a ggplot2 legend to the bottom of the plot.
```{r}
gg + legend_bottom()
```


## Legend to the top
Move a ggplot2 legend to the top of the plot.
```{r}
gg + legend_top()
```


## Legend to the left
Move a ggplot2 legend to the left of the plot.
```{r}
gg + legend_left()
```


## Legend to the right
Move a ggplot2 legend to the right of the plot.
```{r}
gg + legend_right()
```
